---
title: "Exploratory Data Analysis"
knit: (function(input_file, encoding) {
  out_dir <- 'docs';
  rmarkdown::render(input_file,
 encoding=encoding,
 output_file=file.path(dirname(input_file), out_dir, 'index.html'))})
output: html_document
---
Read more about the palmerpenguins package here: https://allisonhorst.github.io/palmerpenguins/

```{r}
# install.packages("palmerpenguins")
library(palmerpenguins)
library(ggplot2)
library(GGally) # Used for pairwise correlation plots
library(dplyr)
library(purrr)
```


Load the palmer penguins data. There are two datasets in this package: penguins and penguins_raw. Penguins has already been pre-cleaned for us, so we will use that
```{r}
data("penguins", package = "palmerpenguins")
```



Take a peek at the data and immediately notice some of the values are NA. Four numeric variables (five if you count year)

```{r}
head(penguins)
```


Looks like there are 11 cases with missing data. Useful to check for any data you're modeling
```{r}
sum(complete.cases(penguins))
```

If you're building linear models, it's important to check for co-linearity among your potential predictors. This will also produce some other useful graphs helping to familiarize us with the data

```{r}
ggpairs(penguins, columns = 1:6)
```

In the bargraph of species x island (second row, first column), its interesting to note that apparently all three species are not found on the same island together, and two species only appear on a single island. 


```{r}
penguins %>% 
   group_by(species, island) %>% 
   summarise(species_count = n())
```
   

Note the correlation between bill length, bill depth, flipper length, and body mass. It appears too that there are two distinct groups present on the bill depth vs flipper length and bill depth vs body mass plots. Let's color these plots by species to see if trends become apparent

```{r}
ggplot(penguins,
       aes(x = bill_depth_mm,
           y = flipper_length_mm,
           colour = species,
           shape = species)) +
   geom_point() +
   theme_minimal() +
   scale_color_manual(values = c("darkorange","purple","cyan4"))

ggplot(penguins,
       aes(x = bill_depth_mm,
           y = body_mass_g,
           colour = species,
           shape = species)) +
   geom_point() +
   theme_minimal() +
   scale_color_manual(values = c("darkorange","purple","cyan4"))
```


Lets look at two more plots with bill length

```{r}
ggplot(penguins,
       aes(x = bill_length_mm,
           y = bill_depth_mm,
           colour = species,
           shape = species)) +
   geom_point() +
   theme_minimal() +
   scale_color_manual(values = c("darkorange","purple","cyan4"))

ggplot(penguins,
       aes(x = bill_length_mm,
           y = flipper_length_mm,
           colour = species,
           shape = species)) +
   geom_point() +
   theme_minimal() +
   scale_color_manual(values = c("darkorange","purple","cyan4"))
```


